---
title: "ST511_Projecr"
output: html_document
---
Introduction:

For this project I have decided to use the IMDB Top 250 Movies database. It contains columns/variables that include Rank, Rating, Title of the movie and the number of Votes.

```{r}
path <- file.path("~", "/Desktop/ST511", "top250.csv" )
#read.csv(path, stringsAsFactors = FALSE)
imdbmovies <- read.csv(path, stringsAsFactors = FALSE)
View(imdbmovies)
```


```{r}
names(imdbmovies)
```

Including an exploratory plot. Here I chose to include a histogram.
```{r}
ggplot(imdbmovies, aes(x=Rating)) + geom_histogram(binwidth=0.1, color="black", alpha=0.9, fill="yellow") + ggtitle("Ratings Histogram") + 
  xlab("Ratings of movies in imdb top 250")
```

```{r}
var(imdbmovies$Rating)
```
We now take a sample out of this data:

```{r}
set.seed(411511)
samp <- sample(1:nrow(imdbmovies), size=50)

imdbmovies[samp, c(3,2)]
```


```{r}

```

**Null hypothesis**: $H_0:\mu=7.5$. In words: the null hypothesis is that the population mean rating is 7.5.

**Alternative hypothesis**: $H_A:\mu>6.5$. In words: the alternative hypothesis is that the population mean IMDB score is greater than 7.5. This is a one-sided greater alternative hypothesis.

We now calculate the Z value, i.e the test statistic:

```{r}
set.seed(411511)
samp <- sample(1:nrow(imdbmovies), size=20)

Xbar <- mean(imdbmovies$Rating[samp])
sigma2 <- var(imdbmovies$Rating)
  
Z <- (Xbar - 7.5) / sqrt(sigma2/20)
Z

Xbar
sigma2
```
```{r}
1-pnorm(Z)
```

**Conclusion**: We fail to reject the null hypothesis $H_0:\mu=7.5$ at significance level $\alpha=0.05$, since our $p$-value is greater than 0.05. We do not have strong evidence that the population mean IMDB Rating is greater than 7.5, so we can not rule out the null hypothesis.


```{r}
sort(c(8,4,5,3,2,1))
```

