---

title: "ST511_Project"
=======
title: "ST511_Projecr"

output: html_document
---
Introduction:

For this project I have decided to use the IMDB Top 250 Movies database. It contains columns/variables that include Rank, Rating, Title of the movie and the number of Votes.

<<<<<<< HEAD
**Population**: 250 films that meet the eligibility criteria
**Variable of Interest**: In this dataset we'll look at the IMDB User Score instead. Possible values are between 0 and 10.
**Scientific question**: Is the average IMDB rating equal to 7.5 for movies in the top 250 list?
**Statistical translation: “Is μ = 7.5 ?”


```{r}
path <- file.path("~", "/Desktop/ST511", "top250.csv" )
#read.csv(path, stringsAsFactors = FALSE)
imdbmovies <- read.csv(path, stringsAsFactors = FALSE)
View(imdbmovies)
```


```{r}

#names(imdbmovies)

names(imdbmovies)

```

Including an exploratory plot. Here I chose to include a histogram.
```{r}

#ggplot(imdbmovies, aes(x=Rating)) + geom_histogram(binwidth=0.1, color="black", alpha=0.9, fill="green") + ggtitle("IMDB Rating") + 
ggplot(imdbmovies, aes(x=Rating)) + geom_histogram(binwidth=0.1, color="black", alpha=0.9, fill="green") + ggtitle("Ratings Histogram") + 

  xlab("Ratings of movies in imdb top 250")
```

```{r}
var(imdbmovies$Rating)
```
<<<<<<< HEAD
**Method**: Here I have chosen to do the t test. 

We now take a sample out of this data:
Let’s randomly sample 20 movies. We will assume that all moviesare equally likely to be selected.


```{r}
#set.seed(411511)
#samp <- sample(1:nrow(imdbmovies), size=10)
#samp1 <- imdbmovies$Rating[samp]
#samp1
samp1 <- c(8.4, 8.2, 8.3, 8.5, 8.0, 8.2, 8.0, 8.4, 8.1, 8.4)
#imdbmovies[samp, c(3,2)]

Xbar <- mean(samp1)
Xbar
sample_variance <- var(samp1)

tstat <- (Xbar-8.2) / sqrt(sample_variance/10)
tstat

 #as n = 20
2*(1-pt(abs(tstat), df=9))
pt

t.test(samp1, mu=8.2)
```

Results: Here, I am stating the hypothesis that has been tested.
=======
We now take a sample out of this data:

```{r}
set.seed(411511)
samp <- sample(1:nrow(imdbmovies), size=50)

imdbmovies[samp, c(3,2)]
```


```{r}

```


**Null hypothesis**: $H_0:\mu=7.5$. In words: the null hypothesis is that the population mean rating is 7.5.

**Alternative hypothesis**: $H_A:\mu>6.5$. In words: the alternative hypothesis is that the population mean IMDB score is greater than 7.5. This is a one-sided greater alternative hypothesis.


**Assesment**: Assess the assumptions/model.

```{r}
#ggplot(imdbmovies, aes(x=Rating)) +
  #geom_histogram(binwidth=50) +
  #facet_wrap(~ Rating)

#hist(imdbmovies)

qqnorm(imdbmovies$Rating)
```



1. The statistical independence of the observations within each sample.
2. The Normality of the underlying population distributions.

To assess assumptions 1 you have to think carefully about how the data were obtained. Typically, data that are collected close together in time and/or space, or data that are collected together in some kind of cluster.

To assess the Normality of the underlying population distributions, you have to take a look at the shapes of the sample distributions and make a judgement call about whether you think the samples could plausibly have come from Normal populations.





We now calculate the Z value, i.e the test statistic:

```{r}
set.seed(411511)
samp <- sample(1:nrow(imdbmovies), size=20)

Xbar <- mean(imdbmovies$Rating[samp])
sigma2 <- var(imdbmovies$Rating)
  
Z <- (Xbar - 7.5) / sqrt(sigma2/20)
Z

Xbar
sigma2
```
```{r}
1-pnorm(Z)
```


**Conclusion**: We fail to reject the null hypothesis $H_0:\mu=7.5$ at significance level $\alpha=0.05$, since our $p$-value is greater than 0.05. We do not have strong evidence that the population mean IMDB Rating is greater than 7.5, so we can not rule out the null hypothesis.


<<<<<<< HEAD


